Internship Summary: Mid-Term Review

### 2020-07-30 RapportStage/Rapport.tex
saving work in sec 4

Simple systems that are analytically solvable (or at least quasi-exactly solvable \titou{(please define)}) are of great importance in theoretical chemistry. 
These systems are very useful to perform benchmark studies in order to test new methods as the mathematics are easier than in realistic systems (such as molecules or solids) but retain much of the key physics. 
To investigate the physics of EPs we consider one such system named \textit{spherium}. 
It consists of two electrons confined to the surface of a sphere interacting through the long-range Coulomb potential. 
Thus, the Hamiltonian is
\begin{equation}
	\hH = -\frac{\grad_1^2 + \grad_2^2}{2} + \frac{1}{r_{12}}
\end{equation}
\titou{T2: I would write the Hamiltonian as a function of the interelectronic angle $\omega$ to highlight the different scaling of the kinetic and potential terms.
This would help following the discussion below which I think is a bit too fast.}

The Laplace operators are the kinetic operators for each electrons and $r_{12}^{-1} = \abs{\vb{r}_1 - \vb{r}_2}^{-1}$ is the Coulomb operator. 
Note that, as readily seen by the definition of the interelectronic distance $r_{12}$, the electrons interact through the sphere.
The radius of the sphere $R$ dictates the correlation regime. 
In the weak correlation regime (i.e., small $R$), the kinetic energy dominates and the electrons are delocalized over the sphere.
For large $R$ (or strong correlation regime), the electron repulsion term drives the physics and the electrons localize on opposite side of the sphere to minimize their Coulomb repulsion. 
This phenomenon is sometimes referred as a Wigner crystalization.
\titou{T2: Missing references in this part.}

@@

In the RHF formalism, the wave function cannot model properly the physics of the system at large $R$ because the spatial orbitals are restricted to be the same. 
If the spatial orbitals are the  same a fortiori it cannot represent two electrons on opposite side of the sphere. 
In the UHF formalism there is a critical value of $R$, called the Coulson-Fischer point \cite{Coulson_1949}, at which a UHF solution appears. 
This solution has broken symmetry as the two electrons tends to localize on opposite side of the sphere. 
\titou{T2: This paragraph could be improved.}

The UHF wave function is defined as
\begin{equation}\label{eq:UHF_WF}
	\Psi_{\text{UHF}}(\theta_1,\theta_2)=\phi_\alpha(\theta_1)\phi_\beta(\theta_2)
\end{equation}
\titou{T2: Please define the RHF wave function first. The angles are not defined.}

Then the one-electron orbitals are expanded in the spatial basis set of the zonal spherical harmonics $Y_{k}(\theta)$:

\begin{equation}
\phi_\alpha(\theta_1)=\sum_{k=0}^{\infty}C_{\alpha,k}\frac{Y_{k0}(\theta_1)}{R}
\end{equation}

It is possible to obtain the formula for the ground state UHF energy in this basis set \cite{Loos_2009}:

@@

\begin{table}[h!]
\centering
\caption{\centering RHF and UHF energies in the minimal basis and exact energies for various R.}
\begin{tabular}{ccccccccc}
\hline
\hline
$R$ & 0.1 & 1 & 2 & 3 & 5 & 10 & 100 & 1000 \\
\hline
RHF 	& 10.00000 & 1.000000 & 0.500000 & 0.333333 & 0.200000 & 0.100000 & 0.010000 & 0.001000 \\
UHF 	& 10.00000 & 1.000000 & 0.490699 & 0.308532 & 0.170833 & 0.078497 & 0.007112 & 0.000703 \\
Exact 	& 9.783874 & 0.852781 & 0.391959 & 0.247898 & 0.139471 & 0.064525 & 0.005487 & 0.000515 \\
\hline
\hline
\end{tabular}
\label{tab:ERHFvsEUHF}
\end{table}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{EsbHF.pdf}
    \caption{\centering Energies of the five solutions of the HF equations (multiplied by $R^2$). The dotted curves correspond to the analytic continuation of the symmetry-broken solutions.}
    \label{fig:SpheriumNrj}
\end{figure}

@@

In this part, we will try to investigate how some parameters of $\hH(\lambda)$ influence the radius of convergence of the perturbation series. The radius of convergence is equal to the distance of the closest singularity to the origin of $E(\lambda)$. Hence we need to determine the locations of the exceptional points to obtain information on the convergence properties. To find them we solve simultaneously the equations \eqref{eq:PolChar} and \eqref{eq:DPolChar}. The equation \eqref{eq:PolChar} is the well-known secular equation giving the energies of the system where $\hI$ is the identity operator. If an energy is also solution of \eqref{eq:DPolChar} then this energy is degenerate. In this case the energies obtained are dependent of $\lambda$ so solving those equations with respect to $E$ and $\lambda$ gives the value of $\lambda$ where two energies are degenerate. These degeneracies can be conical intersections between two states with different symmetry for real value of $\lambda$ or exceptional points between two states with the same symmetry for complex value of $\lambda$.

\begin{equation}\label{eq:PolChar}
\text{det}[E\hI-\hH(\lambda)]=0
\end{equation}

\begin{equation}\label{eq:DPolChar}
\pdv{E}\text{det}[E\hI-\hH(\lambda)]=0
\end{equation}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{EMP_RHF_R10.pdf}
    \caption{\centering Energies $E(\lambda)$ in the restricted basis set \eqref{eq:rhfbasis} with $R=10$.}
    \label{fig:RHFMiniBas}
\end{figure}

@@

The singularity structure in this case is more complex because of the spin contamination of the wave function. We can not use configuration state function in this case. So when we compute all the degeneracies using \eqref{eq:PolChar} and \eqref{eq:DPolChar} some correspond to EPs and some correspond to conical intersections. The numerical distinction of those singularities is very difficult so we will first look at the energies $E(\lambda)$ obtained with this basis set.
Figure \ref{fig:UHFMiniBas} is the analog of Fig.~\ref{fig:RHFMiniBas} in the UHF formalism. We see that in this case the sp\textsubscript{z} triplet interacts with the s\textsuperscript{2} and the p\textsubscript{z}\textsuperscript{2} singlets. Those avoided crossings are due to the spin contamination of the wave function. The exceptional points resulting from those avoided crossings will be discussed in Sec.~\ref{sec:uhfSing}. \\

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{EMP_UHF_R10.pdf}
    \caption{\centering Energies $E(\lambda)$ in the unrestricted basis set \eqref{eq:uhfbasis} with $R=10$.}
    \label{fig:UHFMiniBas}
\end{figure}

>>>>

Simple systems that are analytically solvable (or at least quasi-exactly solvable \titou{(please define)}) are of great importance in theoretical chemistry. 
These systems are very useful to perform benchmark studies in order to test new methods as the mathematics are easier than in realistic systems (such as molecules or solids) but retain much of the key physics. 
To investigate the physics of EPs we consider one such system named \textit{spherium}. 
It consists of two electrons confined to the surface of a sphere interacting through the long-range Coulomb potential. 
Thus, the Hamiltonian is
\begin{equation}
	\hH = -\frac{\grad_1^2 + \grad_2^2}{2} + \frac{1}{r_{12}},
\end{equation}
or
\begin{equation}
	\hH = -\frac{1}{R^2} \qty( \pdv[2]{}{\omega} + \cot \omega \pdv{}{\omega}) + \frac{1}{R \sqrt{2 - 2 \cos \omega}},
\end{equation}
where $\omega$ is the interelectronic angle.
The Laplace operators are the kinetic operators for each electrons and $r_{12}^{-1} = \abs{\vb{r}_1 - \vb{r}_2}^{-1}$ is the Coulomb operator. 
Note that, as readily seen by the definition of the interelectronic distance $r_{12}$, the electrons interact through the sphere.
The radius of the sphere $R$ dictates the correlation regime. 
In the weak correlation regime (i.e., small $R$), the kinetic energy (which scales as $R^{-2}$) dominates and the electrons are delocalized over the sphere.
For large $R$ (or strong correlation regime), the electron repulsion term (which scales as $R^{-1}$) drives the physics and the electrons localize on opposite side of the sphere to minimize their Coulomb repulsion. 
This phenomenon is sometimes referred to as a Wigner crystallization.
\titou{T2: Missing references in this part.}

@@

In the RHF formalism, the two electrons are restricted to ``live'' in the same spatial orbital.
The spatial part of the RHF wave function is then
\begin{equation}\label{eq:RHF_WF}
	\Psi_{\text{RHF}}(\theta_1,\theta_2) = Y_0(\theta_1) Y_0(\theta_2)
\end{equation}
where $\theta_i$ is the polar angle of the $i$th electron and $Y_{\ell}(\theta)$ is a zonal spherical harmonic.

The RHF wave function cannot model properly the physics of the system at large $R$ because the spatial orbitals are restricted to be the same, and, \textit{a fortiori}, it cannot represent two electrons on opposite side of the sphere. 
In the UHF formalism there is a critical value of $R$, called Coulson-Fischer point \cite{Coulson_1949}, at which a UHF solution appears and is lower in energy than the RHF one.
The UHF solution has broken symmetry as the two electrons tends to localize on opposite sides of the sphere. 
The spatial part of the UHF wave function is defined as
\begin{equation}\label{eq:UHF_WF}
	\Psi_{\text{UHF}}(\theta_1,\theta_2)=\phi_\alpha(\theta_1)\phi_\beta(\theta_2)
\end{equation}
where $\phi_\sigma(\theta)$ is the spatial orbital associated with the spin-$\sigma$ electrons ($\sigma = \alpha$ for spin-up electrons and $\sigma = \beta$ for spin-down electrons).
These one-electron orbitals are expanded in the basis of zonal spherical harmonics
\begin{equation}
	\phi_\sigma(\theta)=\sum_{\ell=0}^{\infty}C_{\sigma,\ell}Y_{\ell}(\theta)
\end{equation}
It is possible to obtain the formula for the ground state UHF energy in this basis set \cite{Loos_2009}:

@@

\begin{table}[h!]
\centering
\caption{\centering RHF and UHF energies in the minimal basis and exact energies for various $R$.}
\begin{tabular}{ccccccccc}
\hline
\hline
$R$ & 0.1 & 1 & 2 & 3 & 5 & 10 & 100 & 1000 \\
\hline
RHF 	& 10.00000 & 1.000000 & 0.500000 & 0.333333 & 0.200000 & 0.100000 & 0.010000 & 0.001000 \\
UHF 	& 10.00000 & 1.000000 & 0.490699 & 0.308532 & 0.170833 & 0.078497 & 0.007112 & 0.000703 \\
Exact 	& 9.783874 & 0.852781 & 0.391959 & 0.247898 & 0.139471 & 0.064525 & 0.005487 & 0.000515 \\
\hline
\hline
\end{tabular}
\label{tab:ERHFvsEUHF}
\end{table}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{EsbHF.pdf}
    \caption{Energies of the five solutions of the HF equations (multiplied by $R^2$). The dotted curves correspond to the analytic continuation of the symmetry-broken solutions.}
    \label{fig:SpheriumNrj}
\end{wrapfigure}

@@

In this part, we will try to investigate how some parameters of $\hH(\lambda)$ influence the radius of convergence of the perturbation series. The radius of convergence is equal to the distance of the closest singularity to the origin of $E(\lambda)$. Hence we need to determine the locations of the exceptional points to obtain information on the convergence properties. To find them we solve simultaneously Eqs.~\eqref{eq:PolChar} and \eqref{eq:DPolChar}. Equation \eqref{eq:PolChar} is the well-known secular equation giving the energies of the system where $\hI$ is the identity operator. If an energy is also solution of Eq.~\eqref{eq:DPolChar} then this energy is degenerate. In this case the energies obtained are dependent of $\lambda$ so solving those equations with respect to $E$ and $\lambda$ gives the value of $\lambda$ where two energies are degenerate. These degeneracies can be conical intersections between two states with different symmetry for real value of $\lambda$ or exceptional points between two states with the same symmetry for complex value of $\lambda$.

\begin{subequations}
\begin{align}
	\label{eq:PolChar}
	\text{det}[E\hI-\hH(\lambda)] & = 0,
	\\ 
	\label{eq:DPolChar}
	\pdv{E}\text{det}[E\hI-\hH(\lambda)] & = 0.
\end{align}
\end{subequations}

@@

\begin{wrapfigure}{c}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{EMP_RHF_R10.pdf}
    \caption{Energies $E(\lambda)$ in the restricted basis set \eqref{eq:rhfbasis} with $R=10$.}
    \label{fig:RHFMiniBas}
\end{wrapfigure}

@@

The singularity structure in this case is more complex because of the spin contamination of the wave function. We can not use configuration state function in this case. So when we compute all the degeneracies using Eqs.~\eqref{eq:PolChar} and \eqref{eq:DPolChar} some correspond to EPs and some correspond to conical intersections. The numerical distinction of those singularities is very difficult so we will first look at the energies $E(\lambda)$ obtained with this basis set.
Figure \ref{fig:UHFMiniBas} is the analog of Fig.~\ref{fig:RHFMiniBas} in the UHF formalism. We see that in this case the sp\textsubscript{z} triplet interacts with the s\textsuperscript{2} and the p\textsubscript{z}\textsuperscript{2} singlets. Those avoided crossings are due to the spin contamination of the wave function. The exceptional points resulting from those avoided crossings will be discussed in Sec.~\ref{sec:uhfSing}. \\

\begin{wrapfigure}{c}{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{EMP_UHF_R10.pdf}
    \caption{Energies $E(\lambda)$ in the unrestricted basis set \eqref{eq:uhfbasis} with $R=10$.}
    \label{fig:UHFMiniBas}
\end{wrapfigure}
